%% DaCaF -- Software Impacts: Original Software Publication (OSP)
%% Built on the OFFICIAL "Software Impacts article template" (Version 2, July 2021),
%% the article-class template the journal requires for all submissions.
%%
%% Companion to: Nguyen, Hoang, Hoang, Huynh, "Probabilistic multi-label
%% classification via a divide-and-conquer and fusion approach",
%% Information Fusion (2026), DOI 10.1016/j.inffus.2026.104517
%%
%% Compile: pdflatex dacaf_software_impacts

\documentclass[11pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.5cm]{geometry}
\usepackage{titlesec}
\usepackage{tabu}
\usepackage{enumitem}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{xcolor}
\newlist{selectlist}{itemize}{2}
\setlist[selectlist]{label=$\square$,leftmargin=*,noitemsep,topsep=0pt}

\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{listings}
\usepackage[authoryear,longnamesfirst]{natbib}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
}

\urlstyle{same}

% Set up the section label formatting
\titleformat{\section}[block]{\hspace{1em}\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}[block]{\hspace{1em}}{\thesubsection}{0.5em}{}

\begin{document}

%% ======================================================================
%% Title / name of your software
%% ======================================================================
\noindent
{\Large \textbf{DaCaF: Bayes-optimal per-metric inference for probabilistic
multi-label classification}}
\vskip0.5cm

%% ----- Names of authors / main developers (incl. affiliations, addresses, email)
\noindent
\textbf{Vu-Linh Nguyen}$^{a}$, \textbf{Xuan-Truong Hoang}$^{b,*}$,
\textbf{Anh Hoang}$^{c}$, \textbf{Van-Nam Huynh}$^{b}$\\[2pt]
\noindent
$^{a}$ \textit{Heudiasyc Laboratory, University of Technology of Compi\`egne, Compi\`egne, France}\\
$^{b}$ \textit{Japan Advanced Institute of Science and Technology (JAIST), Nomi, Ishikawa, Japan}\\
$^{c}$ \textit{University of Economics Ho Chi Minh City, Ho Chi Minh City, Vietnam}\\[2pt]
\noindent
$^{*}$ Corresponding author. E-mail: \href{mailto:hxtruong@jaist.ac.jp}{hxtruong@jaist.ac.jp}\\
\vskip0.5cm

%% ======================================================================
%% Abstract (ca. 100 words per the template; trim if required)
%% ======================================================================
\noindent
\textbf{Abstract}\\
In multi-label classification (MLC), different evaluation metrics call for different
Bayes-Optimal Predictions (BOPs), and no single model is optimal for all of them.
Probabilistic MLC allows the target metric to be optimized during the prediction phase,
but computing the exact BOP is intractable in general, and existing tools are
limited, covering only a few metrics, relying on strong assumptions such as label
independence, or lacking a reusable, well-tested implementation. We present DaCaF, a
Python package implementing the Divide-and-Conquer and Fusion approach that, given any
probabilistic model $P(y\mid x)$, returns the exact BOP for a chosen metric, covering two
families of MLC metrics. The package provides seven per-metric prediction rules, each
verified against brute-force enumeration, together with a command-line interface, a
library API, and reproducible evaluation scripts.
\vskip0.5cm

%% ======================================================================
%% Keywords (maximum of six)
%% ======================================================================
\noindent
\textbf{Keywords}\\
Multi-label classification; Probabilistic classifier chains; Bayes-optimal inference;
Evaluation metrics
\vskip0.5cm
\newpage

%% ======================================================================
%% Code metadata (mandatory) -- C1-C9, three columns
%% ======================================================================
\noindent
\textbf{Code metadata}\\

\noindent
\begin{tabular}{|l|p{6.5cm}|p{9.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Please fill in this column} \\
\hline
C1 & Current code version & v1.1.0 \\
\hline
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/hxtruong6/inference_probabilistic_mlc} \newline
Zenodo: \url{https://doi.org/10.5281/zenodo.20572638} \newline
PyPI: \url{https://pypi.org/project/dacaf-mlc/} \\
\hline
C3 & Permanent link to Reproducible Capsule &
\url{https://codeocean.com/capsule/1580907/tree} \\
\hline
C4 & Legal Code License & MIT License \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used &
Python ($\geq$3.10); NumPy, SciPy, scikit-learn, pandas, joblib;
optional PyTorch / torchvision / torchxrayvision ([image] extra) \\
\hline
C7 & Compilation requirements, operating environments \& dependencies &
\texttt{pip install dacaf-mlc} (core) or \texttt{"dacaf-mlc[image]"};
Linux/macOS; Docker image provided \\
\hline
C8 & If available Link to developer documentation/manual &
README.md, CONVENTIONS.md, docs/ \\
\hline
C9 & Support email for questions & hxtruong@jaist.ac.jp \\
\hline
\end{tabular}\\
\vskip0.5cm

%% ======================================================================
%% Body of the article (up to three pages of text)
%% ======================================================================
\section{Motivation and significance}\label{sec:motivation}

Multi-label classification (MLC) is a setting in which each instance can belong to
any subset of the labels, with applications ranging from text and image annotation to
high-stakes domains such as clinical decision-making. To compare ground-truth label sets
and predictions, a variety of evaluation metrics with different characteristics have been
proposed. Probabilistic MLC allows different metrics to be optimized during the prediction
phase: once a probabilistic classifier estimates
the conditional probability of each labeling, the Bayes-Optimal Prediction (BOP) of a
target metric (the prediction $\hat{y}$ that maximizes the expected score of that
metric) can be sought. As different metrics may call for significantly different BOPs, a
mismatch between the target metric, i.e., the metric optimized during the prediction
phase, and the evaluation metric, i.e., the metric used to assess the classifier, can
negatively impact predictive performance.

However, computing the BOP in a naive way, by searching over the $2^L$ candidate
predictions for $L$ labels, quickly becomes intractable as the number of labels grows.
Existing studies are arguably limited in two respects: some focus on only a small number
of metrics, such as subset 0/1 accuracy, Hamming accuracy, and $F_\beta$
\citep{pcc2010,fmax2011,waegeman2014}, yet require a different, metric-specific algorithm
for each metric; others rely on strong assumptions, such as label independence
\citep{pillai2017}. As a consequence, no reusable, exact, and broadly applicable tool
exists for finding BOPs across the metrics commonly used in probabilistic MLC.

We present DaCaF, a software package that implements the Divide-and-Conquer and Fusion
(DaCaF) approach \citep{dacaf2026}: given any probabilistic model $P(y\mid x)$, it returns
the exact BOP for a chosen metric, covering two families of commonly used MLC metrics. It
proceeds in two steps, a divide-and-conquer strategy followed by a fusion strategy, to
find the prediction that is optimal for the selected metric. DaCaF thereby (a) optimizes
the metric one actually cares about, (b) enables studying the mismatch between the target
and evaluation metrics in an exact way, undisturbed by approximation, and (c) provides a
diagnostic that flags metrics whose BOP has a trivial or very restricted
structure, which is useful for assessing the practical value of emerging MLC metrics.

\section{Software description}\label{sec:description}

\subsection{Software architecture}
DaCaF is distributed as the Python package \texttt{dacaf\_mlc}, which can be installed
from PyPI or run through the provided Docker image, and which also exposes a command-line
tool \texttt{dacaf-mlc}. At its core is the class \texttt{ProbabilisticClassifierChainCustom}
(PCC), which turns any scikit-learn base estimator into a probabilistic multi-label
classifier: it arranges the labels as a classifier chain and, from that chain, estimates
the conditional probability $P(y\mid x)$ of each labeling.

Every prediction rule in the package is built from the same two steps of the DaCaF approach.
The \textbf{divide-and-conquer} step uses a property of Bayes-optimal predictions to
partition the $2^L$ possible predictions into $L{+}1$ groups according to the number of
relevant labels; within each group a local optimum can be found efficiently by sorting the
labels by an appropriate score, and the global BOP is then obtained by comparing the local
optima across groups. The \textbf{fusion} step supplies the probabilistic information that
these scores require, namely marginal and pairwise label probabilities, by fusing the predictions
of the dependent binary classifiers that make up the chain, rather than by enumerating all
labelings.

Around this core, the package provides supporting modules for evaluation metrics and their
registry, a pipeline for training and $k$-fold evaluation, the \texttt{evaluate} command-line
interface, and loaders for MULAN ARFF datasets and the NIH ChestX-ray data. A trimmed copy of
the \texttt{skmultiflow} classifier-chain base is vendored so that the package installs and
runs on its own.

\subsection{Software functionalities}
The central functionality is a set of seven per-metric prediction rules (\texttt{predict\_fmeasure},
\texttt{predict\_hamming}, \texttt{predict\_markedness}, \texttt{predict\_precision},
\texttt{predict\_npv}, \texttt{predict\_recall}, and \texttt{predict\_subset}), each of which
returns the prediction that maximizes the expected value of its target metric. Between them
these rules cover two families of MLC metrics, so a single fitted model can be queried for the
Bayes-optimal prediction of many different metrics. The same machinery also yields a diagnostic
that flags metrics whose optimal prediction is trivial or very restricted, which can help
practitioners judge the practical value of a newly proposed metric.

For efficiency, the rules share a batched predictor that issues one \texttt{predict\_proba} call
per chain level instead of evaluating all $N\cdot L\cdot 2^L$ labelings, and its output has been
verified to match brute-force enumeration numerically on small problems. Finally, the package
ships a reproducible evaluation layer (the command-line interface together with scripts over
MULAN tabular datasets and the NIH ChestX-ray data) and a correctness harness that checks every
rule against exact expected-metric enumeration, run under continuous integration on Python 3.10
and 3.12.

\section{Illustrative examples}\label{sec:examples}
A few lines are enough to fit a model once and then query the Bayes-optimal prediction
for any metric of interest:
\begin{lstlisting}[language=Python,basicstyle=\scriptsize\ttfamily]
from sklearn.linear_model import LogisticRegression
from dacaf_mlc.probability_classifier_chains \
    import ProbabilisticClassifierChainCustom
from dacaf_mlc.evaluation_metrics import EvaluationMetrics as EM

pcc = ProbabilisticClassifierChainCustom(
    LogisticRegression(max_iter=10_000))
pcc.fit(X_train, Y_train)            # Y: (n, L) binary
y_f1  = pcc.predict_fmeasure(X_test, beta=1)  # BOP for F1
y_ham = pcc.predict_hamming(X_test)           # BOP for Hamming
print(EM.f_beta(Y_test, y_f1), EM.hamming(Y_test, y_ham))
\end{lstlisting}
Table~\ref{tab:chd49} reports the headline experiment on the CHD-49 dataset, with PCC and
logistic regression averaged over five seeds and ten-fold cross-validation. Each row is the
metric that DaCaF optimizes during prediction, and each column the metric used to evaluate
the resulting predictions; the bold entry on the diagonal marks the largest value in its
column. The pattern is consistent with the divide-and-conquer guarantee: for almost every
evaluation metric, the best score in a column is obtained by optimizing that same metric, so
a mismatch between the target and the evaluation metric measurably lowers performance. The
two exceptions are informative rather than accidental. The NPV and Recall columns are both
maximized by predictions that collapse to the predict-all labeling (the trivial optimum that
the diagnostic flags), which reaches a perfect or near-perfect score without learning anything
metric-specific from the data, while driving the Subset score to zero. The whole table is
regenerated from the shipped code with a single command, \texttt{make reproduce} (CHD-49, the
fastest dataset), and the same API carries over unchanged to the NIH ChestX-ray features
extracted with a pretrained DenseNet or ResNet backbone.

\begin{table}[t]
\centering
\caption{CHD-49, PCC + logistic regression, mean over 5 seeds $\times$ 10-fold CV
(\%, higher is better). Read each column: the bold diagonal entry (optimize the
metric you evaluate) is the maximum of its column. NPV and Recall collapse to the
predict-all trivial optimum.}
\label{tab:chd49}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lccccccc@{}}
\toprule
Target $\downarrow$ / Eval $\rightarrow$ & $F_1$ & Hamming & Markedness & NPV & Precision & Recall & Subset \\
\midrule
Predict $F_1$       & \textbf{67.07} & 69.29 & 71.91 & 81.09 & 62.74 & 79.22 & 15.06 \\
Predict Hamming     & 63.92 & \textbf{70.78} & 71.27 & 75.74 & 66.62 & 66.93 & 18.05 \\
Predict Markedness  & 34.20 & 63.70 & \textbf{76.96} & 71.10 & 33.37 & 40.45 & 8.39 \\
Predict NPV         & 58.35 & 43.01 & 71.50 & \textbf{100.00} & 43.01 & 99.46 & 0.00 \\
Predict Precision   & 40.54 & 64.83 & 68.31 & 63.10 & \textbf{73.52} & 29.10 & 3.10 \\
Predict Recall      & 58.35 & 43.01 & 71.50 & 100.00 & 43.01 & \textbf{99.46} & 0.00 \\
Predict Subset      & 63.96 & 69.37 & 70.35 & 76.47 & 64.15 & 69.79 & \textbf{18.92} \\
\bottomrule
\end{tabular}
\end{table}

\section{Impact}\label{sec:impact}
DaCaF quantifies the exact cost of optimizing a metric other than the one used for
evaluation. Because the package is model-agnostic (a drop-in BOP layer on
top of any probabilistic MLC model that estimates $P(y\mid x)$), this question can be
studied across many estimators and datasets without changing the inference code, and the
answer is exact because no label-independence or sampling approximation gets in the way.
Coverage is wide: a single package handles two whole families of MLC metrics, and its
trivial-optimum diagnostic supports metric selection by flagging metrics whose BOP
collapses to a degenerate form, as NPV and Recall do in Table~\ref{tab:chd49}.

The software is built for reuse beyond the companion paper. Every per-metric rule is checked
against brute-force enumeration and re-run under continuous integration, so the results can
be trusted and rebuilt, and a registry pattern makes adding a new metric or dataset
straightforward while the vendored base estimator keeps the install self-contained.
Getting started takes little: the package is pip-installable and Dockerized, ships scripted
reproduction with bundled MULAN datasets, and lets a newcomer reproduce a paper table in
minutes. We expect the main beneficiaries to be MLC researchers studying how evaluation
metrics behave, practitioners choosing a metric to optimize for deployment, and educators
teaching Bayes-optimal decision-making; an NIH ChestX-ray14 demonstration shows that the
same API also works for clinical and image MLC. The permissive MIT license places no
restriction on this reuse, so the package can also be integrated into proprietary software
and applied in commercial settings.

\section{Conclusions}\label{sec:conclusions}
DaCaF turns a proven theoretical recipe into a reusable and well-tested tool that returns
the exact Bayes-optimal prediction for a chosen metric, given any probabilistic MLC model.
It lets researchers and practitioners optimize the metric they actually care about and study
the cost of a mismatch between the target and evaluation metrics without approximation. Future work includes
supporting more base models and metric families, and adding optional approximate fast paths
for very large label spaces.

%% ======================================================================
%% Required declarations (Guide for Authors)
%% ======================================================================
\vskip0.3cm
\noindent\textbf{Declaration of competing interest}\\
The authors declare that they have no known competing financial interests or personal
relationships that could have appeared to influence the work reported in this paper.
\vskip0.3cm

\noindent\textbf{CRediT authorship contribution statement}\\
%% TODO: confirm each author's CRediT roles (and the final author list) before submission.
%% Suggested (verify with the team):
%%   Vu-Linh Nguyen: Conceptualization, Methodology, Writing -- review & editing.
%%   Xuan-Truong Hoang: Software, Methodology, Validation, Writing -- original draft.
%%   Anh Hoang: Software, Investigation, Data curation.
%%   Van-Nam Huynh: Supervision, Funding acquisition, Writing -- review & editing.
\textbf{Vu-Linh Nguyen}: TODO. \textbf{Xuan-Truong Hoang}: TODO.
\textbf{Anh Hoang}: TODO. \textbf{Van-Nam Huynh}: TODO.
\vskip0.3cm

\noindent\textbf{Acknowledgements}\\
This work was funded/supported by the Junior Professor Chair in Trustworthy AI
(Ref.~ANR-R311CHD) and University of Economics Ho Chi Minh City (UEH), Vietnam. This work
was also sponsored by the Office of Naval Research (ONR) and the Office of Naval Research
Global (ONRG) under grant number N62909-23-1-2058. The views and conclusions contained
herein are those of the authors only and should not be interpreted as representing those
of the U.S. Government.
\vskip0.5cm

%% ======================================================================
%% References
%% ======================================================================
\noindent
\textbf{References}\\
\begin{thebibliography}{9}
\bibitem[Nguyen et~al.(2026)]{dacaf2026} V.-L. Nguyen, X.-T. Hoang, A. Hoang, V.-N. Huynh,
  Probabilistic multi-label classification via a divide-and-conquer and fusion
  approach, Information Fusion (2026), Article 104517.
  \url{https://doi.org/10.1016/j.inffus.2026.104517}.
\bibitem[Hoang et~al.(2026)]{software} X.-T. Hoang, V.-L. Nguyen, A. Hoang, V.-N. Huynh,
  DaCaF: probabilistic multi-label classification (v1.1.0) [Computer software],
  Zenodo (2026). \url{https://doi.org/10.5281/zenodo.20572638}.
\bibitem[Dembczy\'nski et~al.(2010)]{pcc2010} K. Dembczy\'nski, W. Cheng, E. H\"ullermeier,
  Bayes optimal multilabel classification via probabilistic classifier chains, in:
  Proceedings of the 27th International Conference on Machine Learning (ICML), 2010,
  pp. 279--286.
\bibitem[Dembczy\'nski et~al.(2011)]{fmax2011} K. Dembczy\'nski, W. Waegeman, W. Cheng, E. H\"ullermeier,
  An exact algorithm for F-measure maximization, in: Advances in Neural Information
  Processing Systems, Vol. 24, Curran Associates, Inc., 2011.
  \url{https://proceedings.neurips.cc/paper_files/paper/2011/file/71ad16ad2c4d81f348082ff6c4b20768-Paper.pdf}.
\bibitem[Waegeman et~al.(2014)]{waegeman2014} W. Waegeman, K. Dembczy\'nski, A. Jachnik, W. Cheng, E. H\"ullermeier,
  On the Bayes-optimality of F-measure maximizers, J. Mach. Learn. Res. 15 (2014) 3513--3568.
\bibitem[Pillai et~al.(2017)]{pillai2017} I. Pillai, G. Fumera, F. Roli,
  Designing multi-label classifiers that maximize F measures: state of the art,
  Pattern Recognition 61 (2017) 394--404.
\end{thebibliography}

\end{document}
